Aligning marine species range data to better serve science and conservation

### 2016-03-03 data_explore/data_explore.Rmd
updated manuscript

>>>>

# Years for IUCN maps?

``` {r get_iucn_citation_info, echo = FALSE}

  ### Queries shapefiles and extracts column headers.  Use to help identify which
  ### species groups have no id_no field, and possibly other fields to help 
  ### differentiate.

dir_iucn_shp <- file.path(dir_N, 'git-annex/globalprep/_raw_data/iucn_spp/iucn_shp')
groups_list <- as.data.frame(list.files(dir_iucn_shp)) %>%
  rename(shp_fn = `list.files(dir_iucn_shp)`) %>%
  filter(tools::file_ext(shp_fn) == 'shp') %>%
  mutate(shp_fn = str_replace(shp_fn, '.shp', ''))

iucn_shp_info <- data.frame()

for (spp_gp in groups_list$shp_fn) { # spp_gp <- groups_list$shp_fn[3]
  spp_dbf <- foreign::read.dbf(file.path(dir_iucn_shp, sprintf('%s.dbf', spp_gp)))
  message('file read successfully... ')
  cites <- spp_dbf %>%
    select(id_no, compiler, year, citation) %>%
    unique() %>%
    mutate(spp_group = spp_gp)

  iucn_shp_info <- bind_rows(iucn_shp_info, cites)
  message('binding to list...\n')
}

### oldest records go back to late 70s for terrestrial mammals; then skip to 2004 for a marine turtle map.

### 2016-03-03 manuscript-main.Rmd
updated manuscript

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>"
output:
  pdf_document:
    number_sections: yes
    toc: no
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
  word_document: default
---

>>>>

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>"
output:
  word_document: default
  pdf_document:
    number_sections: yes
    toc: no
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
---

### 2016-03-03 mpa_analysis/3_mpa_spp_analysis.Rmd
updated manuscript

@@

>>>>

knitr::kable(mpa_all_runs_wide)

@@

|Scenario               |     n|gap  |covered 0-2% |covered 2-5% |covered 5-10% |covered >10% |
|:----------------------|-----:|:----|:------------|:------------|:-------------|:------------|
|am2014 (50%)           | 17344|1.6% |44.6%        |44.6%        |6.8%          |2.4%         |
|am2015 (50%)           | 22885|1.7% |45.7%        |42.7%        |7.2%          |2.7%         |
|am2015 (0%)            | 22889|1.2% |46.5%        |44.5%        |6.4%          |1.5%         |
|am2015 (0%) + iucn2015 | 24634|2%   |48%          |42.2%        |6.2%          |1.6%         |
|iucn2015               |  4024|6.4% |66.8%        |20.1%        |3.8%          |2.9%         |

### 2016-03-03 ms_2_intro_res_disc.Rmd
finished MPA analysis(?)
updated manuscript

@@

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of species.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique species).  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align, determine several issues that lead to misalignment between predicted species distributions, and outline possible improvements. <!--BSH: I think Jamie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?  CCO: What about age - AquaMaps is pretty recent; before that, would IUCN have been the dominant dataset for marine?  Maybe we can track citations of AquaMaps, but citations of IUCN I think are generic - just cite the Red List - not referencing spatial information specifically, much less marine.-->

We then subject two recent marine biodiversity studies - an assessment of the status of global biodiversity within the Ocean Health Index (Halpern et al. 2012, 2015) and a global analysis of gaps in protection afforded by marine protected areas (MPAs) (Klein et al. 2015) - to a sensitivity analysis. ... ___substituting___ one dataset over the other, to highlight the possible consequences of different data use decisions on our understanding of _the status of_ marine biodiversity. ___BSH suggestion: We used all unique species range data from both sources but substituted one dataset over the other for shared species to highlight the possible consequences of different data use decisions on our understanding of the status of and protection of marine biodiversity.___

@@

The methodologies behind these datasets imply differences in prediction of species distribution due to errors of commission (falsely indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps generally include large commission errors, while predicted distribution models such as AquaMaps will likely include fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals (Rondinini, 2006). 
<!-- This study basically predicted some of the problems we’re seeing here and predicts implications; I’d like to meaningfully acknowledge it, but not sure how deeply to get into it.  Later comments ask about whether to include the omission/commission terminology.

@@

The negative linear relationship <!--BSH: Any stats for this? If so, insert just after this phrase. Otherwise maybe phrase it as 'The general pattern that emerges of a negative linear relationship, where increasing similarity...'--> indicates that increasing similarity in extent correlates with decreasing distribution alignment. <!--BSH: Could also be said as increasing similarity in alignment leads to decreasing similarity in extent.  The phrasing implies causality, even though you are careful to use ‘correlates’--> AquaMaps tends to extrapolate species ranges beyond known occurrences, such that the marginal range predicted by AquaMaps will fall in different locations than the marginal range predicted using IUCN methodology. For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar extents, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

Dividing the map-paired species into quadrants highlights different categories of relationships that in turn help further explain the general pattern. The upper right quadrant includes species (n = `r n_q1`) whose described ranges agree in both spatial distribution and extent. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 2B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because they exist nearly everywhere.

The extent-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their preferred depths, likely introducing errors of commission. <!-- BSH: Worth a few spot checks to confirm this is more or less true. And then maybe include 2-3 examples as figures in the SOM and refer to them here.-->
Ocean depth is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps. _BSH: Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches._ <!--BSH: This is kinda redundant, just emphasizing a solution to the problem.  Not sure it's needed, or perhaps could be woven in to previous sentence.-->

@@

<!--The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor extent match.  Often these mismatches arise when AquaMaps relies on limited data to predict species presence well beyond known occurrences.  Indeed, `r p_q4_datapoor`% (n = `r n_q4_datapoor`) of these cases are data-poor (in which the AquaMaps environmental envelope model is based on fewer than 10 known cell occurrences), compared with `r p_q123_datapoor`% for species in the other three quadrants.  Only `r p_q4_rev`% (n = `r n_q4_rev`) of AquaMaps maps in this quadrant have undergone expert review, compared with `r p_q123_rev`% of species in the other three quadrants. <!--BSH: This implies the problem is solely with AM.  Are there no issues with IUCN maps that cause problems in this quadrant?--> <!--___The data-poor condition likely applies to IUCN as well - AM using OBIS points to generate models, and IUCN is probably doing the same.  Check, and rework.  This means AM is creating errors of commission, while IUCN may be committing errors of omission for poorly-studied species and/or regions - compounding the total problem.___-->

The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor extent match well. Data-poor species are more common in this quadrant; indeed, the median number of species occurrence records (summing (averaging?) OBIS and GBIF occurrences) for this quadrant is `r dp_med_am_q4`, compared to a median of `r dp_med_am_q123` for species in the other three quadrants.  When extrapolating from limited observations, the AquaMaps model often predicts species presence well beyond known occurrences, introducing commission errors; at the same time, IUCN range maps generally target known occurrences, possibly introducing omission errors for data-limited species.

@@

_brief intro to implications - where/how long have IUCN/AquaMaps been used for various purposes; then introduce two case studies_

* ___From JSL:___ _I think you need a broader discussion of implications. How many times has IUCN/Aquamaps been cited? How long have they been around? What scales are the studies? Then you can muse about how results/interpretations would be different depending on which dataset they used, and reference Fig 3’s._
* _Then, say for example, the OHI uses these datasets together (don’t need to get too much into OHI background I think, maybe they don’t even need to know about the biodiversity goal?), but say that to try to combine them they use a threshold at 40% since others use this threshold (but check out the SOM for analyses about thresholds)_

<!--BSH: I think the first bullet point can/should go in the Intro (very briefly!) - I've noted where in my comments to that section. I do agree there needs to be a tiny bit of explanation of how OHI uses the two datasets.-->

@@

The global Ocean Health Index (OHI) (Halpern et al. 2012, 2015), a composite index comprising ten sustainable benefits provided by a healthy ocean, uses species spatial distribution data and IUCN Red List conservation status to calculate biodiversity status (scored from zero to 100) for each of the world's 221 exclusive economic zones. To maximize the number of represented species, OHI gleans spatial distributions from both IUCN and AquaMaps datasets, prioritizing IUCN data for the subset of species in both sources.  OHI uses a probability threshold of 40% to determine species presence for AquaMaps data.

We recalculated the OHI species status score under several scenarios to observe the impact of toggling the prioritized data set from IUCN to AquaMaps, and toggling the AquaMaps presence threshold from 40% to 0% (<a href = '#fig3'>Fig. 3</a>).  Most country scores varied by fewer than five points across the three scenarios; a few country scores decreased by as many as 12 points, indicating a more dire assessment of species conservation status.

_Since the Ocean Health Index Species subgoal relies on spatial data from both datasets, the impacts of these threshold and preference changes will be somewhat muted.  When IUCN range maps are the preferred data source, only the subset of AquaMaps-only species will be affected by threshold changes; and when AquaMaps is the preferred source, the IUCN-only species will dampen the effect of a threshold change._ <!--CCO: This seems overly specific and I think could probably be cut, but the score changes weren’t earth-shattering – and that’s because we are buffering it by using both datasets
BSH: I would maybe reframe this as 'given that only 2166 species overlap, score changes are not likely to be large for OHI and yet we still see important differences . . . " (remember that at the global scale, 2-3 points is a big deal)-->

@@

Klein et al. (2015) compared the global distribution of species to the global distribution of marine protected areas to assess how well the MPAs protect key species and to identify which species fall through gaps in protection.  The study relied on the AquaMaps database, using a probability threshold of 50% or greater, to determine species presence, and the World Database of Protected Areas to define zones of marine protection.  The study determined that currently defined marine protected areas (IUCN categories I through IV overlapping with marine areas) are woefully inadequate in protecting marine biodiversity; 90.5% of species have less than 5% of their overall range represented within marine protected areas.

___working on the analysis...___

_Predictions: What happens when using a 0% threshold instead of a 50%? What happens when using IUCN rather than AM?_
* _IUCN tends to overestimate extent esp for coastal species?  How much of this area will fall within MPAs and how much will fall outside, proportionally? this affects protection ratio._
* _More area means more species (relatively) will be at least partially included in MPAs so fewer apparent gap species?_
* _Policy implications?  Can also tie this back to the predictions of Rondinini et al re: effects of commission/omission errors_

@@

What can be done to address the differences between these two important marine species distribution datasets?  We have identified several potential causes of commission errors among these datasets.  Understanding these causes points toward a few improvements that may help us better align the predictions of species distributions, increasing their utility for ecology research and conservation actions.

@@

* The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (REF https://www.conservationtraining.org/mod/scorm/player.php?a=1412&currentorg=M5L1_ORG&scoid=6492).  Unfortunately, this recommendation appears to be applied inconsistently.  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; even a simple cutoff at the photosynthetic limit of 200 meters would minimize a likely source of commission errors.
* The AquaMaps model is a powerful tool for quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  AquaMaps encourages experts to review and comment on the predicted species distribution maps; lessons from the small sample of reviewed maps can be generalized, for example to other species in the same taxonomic group, to further refine the predictions.  This is especially important for data-poor species.
* In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently the area restrictions are denoted by FAO Major Fishing Areas (REF find a reference in AquaMaps somewhere); however, allowing area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (REF), would likely have better resolution and predictive power, especially for data-poor species.

@@

>>>>

___BSH:___ _I think Julie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?_

___CCO:___ _What about age - AquaMaps is pretty recent (2006 at earliest); before that, would IUCN have been the dominant dataset for marine?  Maybe we can track citations of AquaMaps, but citations of IUCN are generic - just cite the Red List - not referencing spatial information specifically, much less marine.  Going by citation year field on the maps themselves, the oldest marine record I can find is 2004._

_Four foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; 2013; Ready et al. 2010) have been cited a total of 379 times._

@@

To understand the implications of differences between the AquaMaps and IUCN datasets, we compare how each data source represents the global spatial and taxonomic distribution of species.  Most notably, AquaMaps includes range maps for many more species (currently `r format(n_spp_am, big.mark = ',')` species; `r p_spp_am`% of total), such that most global analyses related to biodiversity to date have used AquaMaps (IUCN range map data exist for only `r format(n_spp_iucn, big.mark = ',')` unique species).  For the `r format(n_spp_both, big.mark = ',')` species (`r p_both_all`% of total) mapped in both datasets, we examine how well the maps align, determine several issues that lead to misalignment between predicted species distributions, and outline possible improvements.

We then reexamine two recent marine biodiversity studies - an assessment of the status of global biodiversity within the Ocean Health Index (Halpern et al. 2012, 2015) and a global analysis of gaps in protection afforded by marine protected areas (MPAs) (Klein et al. 2015) - as case studies to explore the implications of prioritizing one data set over the other.  The results highlight the possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

@@

The methodologies behind these datasets imply differences in prediction of species distribution due to errors of commission (falsely _(erroneously? less accusatory, but an annoying word)_ indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps generally include large commission errors, while predicted distribution models such as AquaMaps will likely include fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range <!-- from Rondinini: Omission errors lead to identifying networks of areas for conservation action that are smaller than required and centred on known species occurrences, thus affecting the comprehensiveness, representativeness and efficiency of selected areas. Commission errors lead to selecting areas not relevant to conservation, thus affecting the representativeness and adequacy of reserve network --> (Rondinini, 2006). 
<!-- This study basically predicted some of the problems we’re seeing here and predicts implications; I’d like to meaningfully acknowledge it, but not sure how deeply to get into it.  Later comments ask about whether to include the omission/commission terminology.

@@

A general negative linear pattern emerges, <!--BSH: Any stats for this? If so, insert just after this phrase. Otherwise maybe phrase it as 'The general pattern that emerges of a negative linear relationship, where increasing similarity...'--> suggesting that increasing similarity in extent correlates with decreasing distribution alignment. <!--BSH: Could also be said as increasing similarity in alignment leads to decreasing similarity in extent.  The phrasing implies causality, even though you are careful to use ‘correlates’--> AquaMaps tends to extrapolate species ranges beyond known occurrences, such that the marginal range predicted by AquaMaps will fall in different locations than the marginal range predicted using IUCN methodology. For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar extents, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

By dividing the map-paired species into quadrants, we highlight different categories of relationships that in turn help further explain the general pattern. The upper right quadrant includes species (n = `r n_q1`) whose described ranges agree in both spatial distribution and extent. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 2B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because they exist nearly everywhere.

The extent-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their preferred depths, likely introducing errors of commission. <!-- BSH: maybe include 2-3 examples as figures in the SOM and refer to them here.-->
Ocean depth is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps.  Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches.

@@

The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor extent match well. Data-poor species are more common in this quadrant; indeed, the median number of species occurrence records (summing (averaging?) OBIS and GBIF occurrences) for this quadrant is `r dp_med_am_q4`, compared to a median of `r dp_med_am_q123` for species across the other three quadrants.  When extrapolating from limited observations, the AquaMaps model often predicts species presence well beyond known occurrences, introducing commission errors; at the same time, IUCN range maps generally target known occurrences, possibly introducing omission errors for data-limited species.

@@

Clear and significant differences in species range descriptions between AquaMaps and IUCN range maps, due to method-driven differences in errors of commission and omission.  To examine the implications of these differences, we replicated two recent studies, varying only the prioritization of one data set over the other.

@@

The global Ocean Health Index (OHI) (Halpern et al. 2012, 2015), a composite index comprising ten sustainable benefits provided by a healthy ocean, uses species spatial distribution data and IUCN Red List conservation status to calculate biodiversity status (scored from zero to 100) for each of the world's 221 exclusive economic zones. To maximize the number of represented species, OHI gleans spatial distributions from both IUCN and AquaMaps datasets, prioritizing IUCN data for the `r n_spp_both` species included in both sources.  OHI currently uses a probability threshold of 40% to determine species presence for AquaMaps data.

We recalculated the OHI species status score under several scenarios to observe the impact of toggling the prioritized data set from IUCN to AquaMaps, and toggling the AquaMaps presence threshold from 40% to 0% (<a href = '#fig3'>Fig. 3</a>).  Given that only `r n_spp_both` species overlap out of `r n_spp_all`, it is surprising that changing the priority for overlapping species from IUCN maps to AquaMaps would result in such large score shifts as seen in scenarios 2 and 3.  While many country scores showed only small variation across all three scenarios, select countries gained as many as 5 points while others dropped up to 12 points.  Significantly, the global score decreased by two full points in scenario 3, indicating that an arbitrary change in how the two datasets are combined results in a more dire assessment of species conservation status.

@@

Klein et al. (2015) compared the global distribution of species to the global distribution of marine protected areas to assess how well the MPAs protect key species and to identify which species fall through gaps in protection.  The study relied on the AquaMaps database, using a probability threshold of 50% or greater, to determine species presence, and the World Database of Protected Areas to define zones of marine protection.  The study determined that currently established MPAs (IUCN categories I through IV overlapping with marine areas) are woefully inadequate in protecting marine biodiversity; 90.5% of species have less than 5% of their overall range represented within MPAs and 1.4% of species have no protection at all ("gap" species).

 (<a href = '#fig4'>Figure 4</a>) displays the results of applying this methodology under several different scenarios.  When comparing the conclusions based upon an IUCN-only scenario to an AquaMaps-only scenario (using 2015 data and a 0% threshold sets the most meaningful comparison) we found a five-fold increase in the proportion of gap species (6.4% vs 1.2%) and dramatically larger proportion of species with less than 2% of their range protected (73.2% vs. 47.7%).  However, this comparison also indicates a larger proportion of well-protected species with greater than 10% of range protected (2.9% vs. 1.5%).  Conclusions drawn from each of these datasets would paint dramatically different pictures of the protection afforded by our current global MPA network.

@@

AquaMaps and IUCN range maps provide important spatial information on marine species ranges, and show reasonable agreement for many well-studied species, but illustrate uncertainty in our understanding of spatial distribution for many others.  Identifying and addressing differences in these datasets will increase their utility for ecology research and conservation actions.  We have identified several likely drivers of commission errors and omission errors between these datasets, pointing toward improvements to better align the predictions of species distributions.

@@

* The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (REF https://www.conservationtraining.org/mod/scorm/player.php?a=1412&currentorg=M5L1_ORG&scoid=6492).  Unfortunately, this recommendation appears to be applied inconsistently.  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; even a simple cutoff at the photosynthetic limit of 200 meters would minimize a likely source of commission errors.
* The AquaMaps model is a powerful tool for quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  AquaMaps encourages experts to review and comment on the predicted species distribution maps; lessons from the small sample of reviewed maps can be generalized, for example to other species in the same taxonomic group, to further refine the predictions.  This is especially important for data-poor species.
* In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently the area restrictions are denoted by FAO Major Fishing Areas (REF find a reference in AquaMaps somewhere); however, area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (REF), would likely provide better resolution and predictive power, especially for data-poor species.

@@

_(do we need a wrap-it-up paragraph here?  I put that more at the top of the conclusions section... could shift it down here, but probably not needed in both places)_

### 2016-03-03 ms_4_figs.Rmd
finished MPA analysis(?)
updated manuscript

### Figure 2

@@

> __Fig. 2.__ (A) Distribution alignment (overlap of smaller range within larger) versus extent ratio (the ratio of smaller range area to the larger range area) for `r n_spp_both` species included in both IUCN and AquaMaps datasets.  The upper right quadrant (quadrant 1) comprises species whose maps largely agree (better than median value) in both spatial distribution and the extent of described ranges
(n = `r q1_count`; `r round(q1_count/nrow(spp_map_pairs) * 100, 1)` %).
> The upper left quadrant (quadrant 2) comprises species whose maps agree well in distribution, but disagree in extent
(n = `r q2_count`; `r round(q2_count/nrow(spp_map_pairs) * 100, 1)` %).
> The lower right quadrant (quadrant 3) includes species for which the paired maps generally agree in range extent, but disagree on where those ranges occur
(n = `r q3_count`; `r round(q3_count/nrow(spp_map_pairs) * 100, 1)` %).
> The lower left quadrant (quadrant 4) indicates species for which the map pairs agree poorly in both area and distribution
(n = `r q4_count`; `r round(q4_count/nrow(spp_map_pairs) * 100, 1)` %).  
(B) Alignment breakdown of paired-map species by taxonomic group. _is it distracting to use order names for some groups and not others?  Some of the IUCN taxonomic groups had pretty long names ("acanthuridae" is way shorter than "surgeonfish, tangs, and unicornfish")_

@@

___NOTE: These figures still need to be recreated using the latest and greatest data!___ _currently they use data generated in November or so, in ohiprep._

> __Fig. 3.__ Change in status score for the Species Subgoal within the global Ocean Health Index under three different scenarios.

| Scenario              | Priority data source | AquaMaps presence threshold |
| --------------------- | :------------------: | :-------------------------: |
| Scenario 0 (current)  | IUCN                 |  >= 40%                     |
| Scenario 1            | IUCN                 |   > 0%                      |
| Scenario 2            | AquaMaps             |  >= 40%                     |
| Scenario 3            | AquaMaps             |   > 0%                      |

> * Scenario 1 shows the effect of reducing the presence threshold for AquaMaps presence.  Reducing the threshold increases the apparent range of a species; the slight decrease in average score suggests increased spatial representation of threatened species.
> * Scenario 2 shows the effect of prioritizing AquaMaps data over IUCN, while maintaining the same presence threshold.  This will have different effects depending on the species; in general, AquaMaps ranges are smaller than IUCN ranges, so most overlapping species will see a decrease in represented range.  The slight bump in mean score may indicate a small increase in spatial representation of low-risk species, a small decrease in spatial representation of high-risk species, or more likely a combination of both.
> * Scenario 3 shows the effect of prioritizing AquaMaps data over IUCN, while eliminating the presence threshold.  Just as a presence threshold of zero in scenario 1 drives a decrease in average score relative to the baseline, the zero threshold in scenario 3 drives a decrease in scores relative to scenario 2.  The large decrease indicates that within the set of paired-map species, a zero threshold greatly increases the spatial representation of high-risk species relative to low-risk species.

>>>>

@@

> __Fig. 2.__ (A) Distribution alignment (overlap of smaller range within larger) versus extent ratio (the ratio of smaller range area to the larger range area) for `r n_spp_both` species included in both IUCN and AquaMaps datasets.  The upper right quadrant (quadrant 1) comprises species whose maps largely agree (better than median value) in both spatial distribution and the extent of described ranges
(n = `r q1_count`; `r round(q1_count/nrow(spp_map_pairs) * 100, 1)` %).
> The upper left quadrant (quadrant 2) comprises species whose maps agree well in distribution, but disagree in extent
(n = `r q2_count`; `r round(q2_count/nrow(spp_map_pairs) * 100, 1)` %).
> The lower right quadrant (quadrant 3) includes species for which the paired maps generally agree in range extent, but disagree on where those ranges occur
(n = `r q3_count`; `r round(q3_count/nrow(spp_map_pairs) * 100, 1)` %).
> The lower left quadrant (quadrant 4) indicates species for which the map pairs agree poorly in both area and distribution
(n = `r q4_count`; `r round(q4_count/nrow(spp_map_pairs) * 100, 1)` %).  
(B) Alignment breakdown of paired-map species by taxonomic group.

@@

> __Fig. 3.__ Change in status score for the Species Subgoal within the global Ocean Health Index under three different scenarios.
> Scenario 1 shows the effect of reducing the probability threshold for AquaMaps presence to more accurately track the definition of IUCN “limits of distribution."  Reducing the threshold increases the apparent range of a species; the slight decrease in average score suggests increased spatial representation of threatened species.
> Scenario 2 shows the effect of prioritizing AquaMaps data over IUCN, while maintaining the 40% presence threshold.  In general, AquaMaps ranges are smaller than IUCN ranges, so most overlapping species will see a decrease in represented range.  The slight bump in mean score may indicate a small increase in spatial representation of low-risk species, a small decrease in spatial representation of high-risk species, or a combination of both.
> Scenario 3 shows the effect of prioritizing AquaMaps data over IUCN, and simultaneously eliminating the presence threshold.  The zero threshold in scenario 3 drives a decrease in scores relative to scenario 2.  The large decrease indicates that a zero threshold greatly increases the spatial representation of high-risk species.

___NOTE: This figure still needs to be recreated using the latest and greatest data!___ _currently they use data generated in November or so, in ohiprep._

### Figure 4

<a id = 'fig4'></a>

![Change in MPA coverage](figures/barchart_mpa_gap.png)

> __Fig. 4.__  Percent of species range covered by MPAs based upon methods in Klein et al. (2015).  Scenario 1 replicates the original results, measuring protected range of species in AquaMaps 2014 dataset, with a 50% presence threshold, against the 2014 World Database of Protected Areas, filtered for IUCN categories I-IV that overlap marine areas.  Scenario 2 updates the results using the 2015 AquaMaps dataset; the small changes despite the inclusion of an additional 5545 species demonstrates the robustness of the conclusions of Klein et al. (2015).  Scenario 3, using 2015 AquaMaps data, drops the presence threshold to zero, showing an expected decrease in gap species, but also a decrease in species with 5% or greater protected range.  Scenario 4 adds an additional 1745 species unique to the IUCN dataset, resulting in increases in gap species and species with less than 2% coverage.  Scenario 5 examines species MPA coverage using only the IUCN dataset.

`r knitr::kable(read_csv(file.path(dir_data, 'mpa_all_runs_summary.csv')))`

### 2016-03-03 ms_5_refs.Rmd
updated manuscript

Martin C.S., Fletcher R., Jones M.C., Kaschner K., Sullivan E., Tittensor D.P., Mcowen C., Geffert J.L., van Bochove J.W., Thomas H., Blyth S., Ravillious C., Tolley M., Stanwell-Smith D. (2014). Manual of marine and coastal datasets of biodiversity importance. May 2014 release. Cambridge (UK): UNEP World Conservation Monitoring Centre. 28 pp. (+ 4 annexes totalling 174 pp. and one e-supplement). PDF _report on marine data sets and data gaps etc, incl both IUCN and AM_

@@

>>>>

Martin C.S., Fletcher R., Jones M.C., Kaschner K., Sullivan E., Tittensor D.P., Mcowen C., Geffert J.L., van Bochove J.W., Thomas H., Blyth S., Ravillious C., Tolley M., Stanwell-Smith D. (2014). Manual of marine and coastal datasets of biodiversity importance. May 2014 release. Cambridge (UK): UNEP World Conservation Monitoring Centre. 28 pp. (+ 4 annexes totalling 174 pp. and one e-supplement). PDF _review of marine data sets and data gaps etc, incl both IUCN and AM as well as many others_

@@

Predicting the distributions of marine organisms at the global scale
Ready, J., Kaschner, K., South, A. B., Eastwood, P. D., Rees, T., Rius, J., ... & Froese, R. (2010). Predicting the distributions of marine organisms at the global scale. Ecological Modelling, 221(3), 467-478.
Cited by 75

[PDF] Mapping world-wide distributions of marine mammal species using a relative environmental suitability (RES) model
K Kaschner, R Watson, AW Trites, D Pauly - Marine Ecology Progress …, 2006 - vliz.be
ABSTRACT: The lack of comprehensive sighting data sets precludes the application of 
standard habitat suitability modeling approaches to predict distributions of the majority of 
marine mammal species on very large scales. As an alternative, we developed an ...
Cited by 167 Related articles All 12 versions Web of Science: 86 Cite Save

[DOC] from aquamaps.org
[DOC] AquaMaps
K Kaschner, J Rius-Barile, K Kesner-Reyes… - Predicted range maps …, 2013 - aquamaps.org
AquaMaps is an approach to generating model-based, large-scale predictions of currently 
known natural occurrence of marine species. Models are constructed from estimates of the 
environmental tolerance of a given species with respect to depth, salinity, temperature, ...
Cited by 64 Related articles All 3 versions Cite Save More

[HTML] Current and future patterns of global marine mammal biodiversity
K Kaschner, DP Tittensor, J Ready, T Gerrodette… - PLoS …, 2011 - journals.plos.org
Abstract Quantifying the spatial distribution of taxa is an important prerequisite for the 
preservation of biodiversity, and can provide a baseline against which to measure the 
impacts of climate change. Here we analyse patterns of marine mammal species richness ...
Cited by 73 Related articles All 20 versions Web of Science: 34 Cite Save More
